---
title: "Data Analysis on Predict test scores of students"
output:
  html_document: default
  word_document: default
---

### Data loading

```{r}
df = read.csv('test_scores.csv')
head(df)
```

### The type of object can be ascertained using the class() command

```{r}
class(df)
```

```{r}
class(df$posttest)
```

```{r}
class(dimnames(df))
```

### Shows the top of the data object and by default shows the first six rows:

```{r}
head(df)
```

### Can elect to show a different number of rows using the n = instruction like so:

```{r}
head(df, n=3)
```

### can also display the bottom of the data using the tail() command default shows the last six rows:

```{r}
tail(df)
```

### Can elect to show a different number of rows using the n = instruction like so:
```{r}
tail(df, n=3)
```

### To get information about an Data frame and particular columns:

```{r}
summary(df)
```

```{r}
summary(df$posttest)
```

```{r}
mean(df$posttest)
```

```{r}
median(df$posttest)
```

### Shows the median absolute deviation

```{r}
mad(df$posttest)
```

```{r}
mode(df$posttest)
```

```{r}
var(df$posttest)
```

```{r}
sd(df$posttest)
```

```{r}
quantile(df$posttest)
```

### Returns Tukey's five number summary (minimum, lower-hinge, median, upper-hinge, maximum) for the input data.

```{r}
fivenum(df$posttest)
```

### F Test to Compare Two Variances

```{r}
var.test(df$posttest, df$pretest)
```

###  str() command is useful to see the object structure

```{r}
str(df)
```

### structure returns the given object with further attributes set

```{r}
head(structure(df))
```

### can also look at all the named objects you have at once using the ls.str() command

```{r}
ls.str(df)
```

### can use the pattern = instruction to narrow down your focus

```{r}
ls.str(pattern = 'df')
```

### The most basic command that enables the viewing of column or row is:

```{r}
names(df)
```

```{r}
row.names(df)
```

```{r}
colnames(df)
```

### dimnames(): this looks at both the row and column names at the same time:

```{r}
dimnames(df)
```

### length() command used to determine the number of items in an object

```{r}
length(df)
```

### To extract a particular column and particular row values

```{r}
head(df$posttest)
```

```{r}
df$posttest[3]
```

```{r}
df$posttest[1:5]
```

```{r}
tdf = attach(df)
tdf
```

### The max() and min() commands display the largest and smallest values in a numeric object

```{r}
max(df$posttest)
```

```{r}
max(df$pretest)
```

```{r}
min(df$posttest)
```

```{r}
min(df$pretest)
```


```{r}
head(stack(df))
```

### Sorting default is ascending order

```{r}
head(sort(df$posttest))
```

```{r}
head(sort(df$pretest))
```

### can get an index using the order() command. This uses the same instructions as the sort() command, but tells you the position of each item along the vector:

```{r}
head(order(df$posttest))
```

### The rank() command gives the rank number like order() in a slightly different manner when the values are same the ranks are shared between them which is not in order()

```{r}
head(rank(df$posttest), n=10)
```

### Displaying selected rows & columns

```{r}
df[3,3]
```

```{r}
df[3, 1:5]
```

```{r}
df[c(1,455,1000),]
```

```{r}
df[c(511,652,121), "school"]
```

```{r}
df[c(65,515,114), 1:5 ]
```

### TO convert from Data frame from matrix

```{r}
head(as.matrix(df))
```

### The cbind() and rbind() commands assemble a matrix/data frames, by columns or rows, from several other object

```{r}
head(rbind(df))
```

```{r}
head(cbind(df))
```

### Simple Cumulative Commands

```{r}
head(cumsum(df$posttest))
```

```{r}
head(cummax(df$pretest))
```

```{r}
head(cummin(df$posttest))
```

### The cumulative product

```{r}
head(cumprod(df$posttest))
```

### The table() command enables you to specify which columns of data you want to use to create your contingency table

```{r}
table(df$posttest)
```

```{r}
head(as.table(df$posttest))
```

### To check whether the data extracted is data frame or not

```{r}
is.data.frame(df)
```

```{r}
is.table(df)
```

### The class() command can form the basis of a logical test by using the if() command in the following manner:

```{r}
if(class(df) == 'data.frame') TRUE else FALSE
```

### The seq_along() command creates a simple index

```{r}
seq_along(df)
```

### Create a basic stem and leaf plot using the stem() command

```{r}
stem(df$posttest)
```

### Now increase the number of bins used by adding a scale = 2 instruction

```{r}
stem(df$posttest, scale = 2)
```

### stem with a conditional statement:

```{r}
with(df, stem(df$posttest[df$school == 'ANKYI']))
```

### Histogram Plotting:

```{r}
hist(df$posttest)
```

```{r}
hist(df$pretest)
```

```{r}
hist(df$n_student)
```

```{r}
hist(df$posttest, col = 'lightblue', xlab = 'PostTest', ylim = c(0, 0.1), freq = FALSE)
```

### Have seen in drawing a histogram with the hist() command that you can use freq = FALSE to force the y-axis to display the density rather than the frequency of the data. You can also call on the density function directly via the density() command.

```{r}
density(df$posttest)
```

### Using the Density Function to Draw a Graph

```{r}
plot(density(df$posttest))
```

```{r}
plot(density(df$pretest))
```

```{r}
hist(df$posttest, freq = F)
lines(density(df$posttest), lty = 2)
lines(density(df$posttest, k = 'rectangular'))
```

### Generates n random numbers from the normal distribution with mean of 0 and standard deviation of 1

```{r}
head(rnorm(df$posttest, mean = 0, sd = 1))
```

### Returns the probability for the quantile q

```{r}
head(pnorm(df$posttest, mean = 0, sd = 1))
```

### Returns the quantile for a given probability p

```{r}
head(qnorm(df$posttest, mean = 0, sd = 1))
```

### Gives the density function for values x

```{r}
head(dnorm(df$posttest, mean = 0, sd = 1))
```

### Quantile-Quantile Plot

```{r}
qqnorm(df$posttest)
```

```{r}
qqnorm(df$pretest)
```

```{r}
qqnorm(df$posttest, main = 'QQ plot of df$posttest', xlab = 'Theoretical', 
ylab = 'Quantiles for df$posttest')
qqline(df$posttest, lwd = 2, lty = 2)
```

### can also plot one distribution against another as a quantile-quantile plot using the qqplot() command.

```{r}
qqplot(df$pretest, df$posttest)
```

### would be useful to draw a straight line on your qqplot() and you can do that using the abline() command. This command uses the properties of a straight line (that is, y = a + bx) to produce a  line on an existing plot. The general form of the command is: abline(a = intercept, b = slope)

###  lm(), which carries out linear modeling. This command determines the line of best fit between the x and y values in your qqp object.

```{r}
qqp = qqplot(df$pretest, df$posttest)
abline(lm(qqp$y ~ qqp$x))
```

### The basic method of applying a t-test is to compare two vectors of numeric data

```{r}
t.test(df$pretest, df$posttest)
```

### can override the default and use the classic t-test by adding the var.equal = TRUE instruction, which forces the command to assume that the variance of the two samples is equal.

```{r}
t.test(df$pretest, df$posttest, var.equal = TRUE)
```

### can also carry out a one-sample t-test. In this version you supply the name of a single vector and the mean to compare it to (this defaults to 0):

```{r}
t.test(df$posttest, mu = 5)
```

### Using Directional Hypotheses

```{r}
t.test(df$posttest, mu = 5, alternative = 'greater')
```

### Formula Syntax and Subsetting Samples in the t-Test

```{r}
t.test(df$posttest ~ df$school, data = df, subset = df$school %in% c('ANKYI' , 'CCAAW'))
```

### Two-Sample U-Test

```{r}
wilcox.test(df$pretest, df$posttest)
```

### Carries out a basic correlation between x and y . If x is a matrix or data frame, y can be omitted

```{r}
cor(df$pretest, df$posttest)
```

### Determines covariance between x and y . If x is a matrix or data frame, y can be omitted 

```{r}
x = cov(df$pretest, df$posttest)
x
```

### The cov2cor() command is used to determine the correlation from a matrix of covariance in the following example:

```{r}
v = as.matrix(x)
cov2cor(v)
```

### Significance Testing in Correlation Tests

```{r}
cor.test(df$pretest, df$posttest)
```

### Chi-squared tests of association can be carried out using the chisq.test() command.

```{r}
chisq.test(df$posttest)
```

### BoxPlot's:

```{r}
boxplot(df$posttest)
```

```{r}
boxplot(df$pretest)
```

```{r}
boxplot(df$pretest, df$posttest)
```

```{r}
boxplot(df$pretest ~ df$school, data = df)
```

```{r}
boxplot(df$posttest ~ df$school, data = df)
```

```{r}
boxplot(df$posttest ~ df$gender, data = df)
```

```{r}
boxplot(df$pretest ~ df$gender, data = df)
```

```{r}
boxplot(df$posttest ~ df$school_setting)
```

```{r}
boxplot(df$posttest ~ df$school_type)
```

```{r}
boxplot(df$posttest ~ df$teaching_method)
```

```{r}
boxplot(df$posttest ~ df$lunch)
```

### Horizontal Boxplots

```{r}
boxplot(df$posttest, horizontal = T)
title(xlab = 'PostTest scores')
```

### scatter plot's
filtering the rows by school level (ANKYI) and storing the df in temp object for ease of scatter plotting and for further usage

```{r}
library(dplyr)
temp = df %>% filter(df$school == "ANKYI")
plot(temp$pretest, temp$posttest, pch = 25)
```

### The pch = instruction refers to the plotting character, and can be specified in one of several ways. You can type an integer value and this code will be reflected in the symbol/character produced. For values from 0 to 25

```{r}
with(df, plot(df$posttest[df$school == 'ANKYI'], pch = 0))
```

### with() command can be used to get the particular values based on row(level) conditions

```{r}
with(df, (df$posttest[df$school == 'ANKYI']))
```

### the pairs() command takes all the columns in a data frame and creates a matrix of scatter plots.

```{r}
pairs(~ pretest + posttest, data = temp)
```

### Line plot

```{r}
plot(temp$posttest, type = 'l' )
```

```{r}
plot(temp$posttest, type = 'b')
```

### Pie chart

```{r}
pie(temp$posttest, labels = 'PostTest score')
```

```{r}
pie(temp$pretest, labels = 'PreTest score')
```

### Cleveland Dot Charts

```{r}
dotchart(temp$posttest)
```

### Bar plot

```{r}
barplot(temp$posttest)
title(xlab = 'ANKYI school', ylab = 'PostTest score')
box()
```

### table() used to get the column's diff level(type) counts

```{r}
table(df$school)
```

```{r}
barplot(temp$posttest, horiz = T)
title(xlab = 'ANKYI school', ylab = 'PostTest score')
box()
```

### the analysis of variance using the aov() command

```{r}
temp.aov = na.omit(aov(temp$pretest ~ temp$posttest, data = temp))
temp.aov
```

### To see the classic ANOVA table of results you need to use the summary() command like so

```{r}
summary(temp.aov)
```

### Two-way ANOVA

```{r}
toway = aov(posttest ~ pretest * n_student, data = temp)
toway
```

```{r}
interaction.plot(temp$posttest, temp$pretest, temp$n_student)
```

```{r}
factor(df$school_setting)
```

### creating new testCol/testRow for addition as a new column/row to the df object (using rep() command)

```{r}
testCol = c(rep(df$posttest - df$pretest, length(df)))
head(testCol)
```

```{r}
testRow = c('NMIT', 'Urban', "Non-public", 263, 'Standard', 18, '1NT20SDS09', 'Male', 'NA', 00, 00)
testRow
```

### Adding Rows or Columns

```{r}
test = data.frame(df, testCol)
head(test)
```

```{r}
test = df
test['testRow',] = testRow
tail(test)
```

```{r}
test = df
test["Test",] = NA
tail(test)
```

### Column Indexes

```{r}
col(temp, as.factor = F)
```

### Extracting two(2) columns from df

```{r}
x = df[ , 10:11]
head(x)
```

### Simple Column and Row Summaries

```{r}
colMeans(x)
```

```{r}
rowMeans(x)
```

```{r}
colSums(x)
```

```{r}
rowSums(x)
```

### The apply() command to apply a function over all the rows or columns of a data frame (or matrix)

Same as colMean() in above code

```{r}
apply(df[ , 10:11], 2, mean)
```

```{r}
apply(df[ , 10:11], 1, mean)
```

```{r}
apply(df[ , 10:11], 2, median)
```

```{r}
apply(df[ , 10:11], 2, var)
```

### Using tapply() to Summarize Using a Grouping Variable

```{r}
tapply(df$posttest, df$school, FUN = mean)
```

```{r}
tapply(df$posttest, df$school, FUN = var)
```

```{r}
tapply(df$posttest, df$school_setting, FUN = mean)
```

```{r}
tapply(df$posttest, df$school_type, FUN = mean)
```

```{r}
tapply(df$posttest, df$teaching_method, FUN = mean)
```

```{r}
tapply(df$posttest, df$gender, FUN = mean)
```

```{r}
tapply(df$posttest, df$lunch, FUN = mean)
```

### The aggregate() command enables you to compute summary statistics for subsets of a data frame or matrix; the result comes out as a single matrix rather than an array item, even with multiple grouping factors

```{r}
aggregate(df[ , 10:11], by = list(df$school), FUN = mean)
```

```{r}
aggregate(df[ , 10:11], by = list(df$school_setting), FUN = mean)
```

```{r}
aggregate(df[ , 10:11], by = list(df$school_type), FUN = mean)
```

```{r}
aggregate(df[ , 10:11], by = list(df$teaching_method), FUN = mean)
```

```{r}
aggregate(df[ , 10:11], by = list(df$gender), FUN = mean)
```

### uding cbind()

```{r}
aggregate(cbind(df$pretest, df$posttest), data = df, by = list(df$lunch), FUN = mean)
```

```{r}
aggregate(df$posttest ~ df$school, data = df, FUN = mean)
```

```{r}
aggregate(df$posttest ~ df$school * df$school_type, data = df, FUN = mean)
```

### The na.omit() command strips out unwanted NA items from vectors and data frames.

```{r}
head(na.omit(df))
```
### ls and objects return a vector of character strings giving the names of the objects in the specified environment

```{r}
objects(df)
```

```{r}
ls(df)
```

### Simple lineaR RegReSSion

```{r}
dflm = lm(posttest ~ school, data = df)
dflm
```

```{r}
cor.test(~ df$pretest + df$posttest, data = df)
```

### Linear Model Results Objects

###  can extract the coefficients using the coef() command

```{r}
coef(dflm)
```

### can obtain confidence intervals on these coefficients using the confint() command

```{r}
confint(dflm)
```

### Fitted Values

```{r}
head(fitted(dflm))
```

### Residuals

```{r}
head(residuals(dflm))
```
### Formula

```{r}
formula(dflm)
```

```{r}
dflm$call
```

```{r}
formula(df)
```

### Using the segments() Command for Error Bars

```{r}
df.m = apply(df[,10:11], 2, mean)
df.m
```

```{r}
df.sd = apply(df[ , 10:11], 2, sd)
df.sd
```

```{r}
df.s = apply(df[,10:11], 2, sum)
df.s
```

```{r}
df.1 = df.s/df.m
df.1
```

```{r}
df.se = df.sd / sqrt(df.1)
df.se
```

```{r}
df.m + df.se
```

```{r}
max(df.m + df.se)
```

```{r}
df.max = round(max(df.m + df.se) + 0.5, 0)
df.max
```

```{r}
bp = barplot(df.m, ylim = c(0, df.max))
bp
```

```{r}
bp = barplot(df.m, ylim = c(0, df.max))
segments(bp, df.m + df.se, bp, df.m - df.se)
segments(bp - 0.2, df.m - df.se, bp + 0.2, df.m - df.se)
box()
title(xlab = 'Type of test', ylab = 'Scores')
```

### Creating Mathematical Expressions

```{r}
plot(1:10, 1:10, type = 'n')
opt = par(cex = 1.5)
text(1, 1, expression(hat(x)))
text(2, 2, expression(alpha==x))
text(3, 3, expression(beta==y))
text(4, 4, expression(frac(x, y)))
text(5, 5, expression(sum(x)))
text(6, 6, expression(sum(x^2)))
text(7, 7, expression(bar(x) == sum(frac(x[i], n), i==1, n)))
text(8, 8, expression(sqrt(x)))
text(9, 9, expression(sqrt(x, 3)))
```

```{r}
plot(temp$pretest ~ temp$posttest, data = temp, pch = 21, ylab = 'PreTest', xlab = 'PostTest')
points(temp$pretest ~ temp$posttest, data = temp, pch = 19)
legend(x = 'topright', legend = c('PreTest', 'PostTest'), pch = c(21,19), bty = 
'n')
```

```{r}
curve(sin, -pi*2, pi*2, lty = 2, lwd = 1.5, ylab = 'Function', ylim = c(-1,1.5))
curve(cos, -pi*2, pi*2, lty = 3, lwd = 1, add = TRUE)
legend(x = 'topright', legend = c('Sine', 'Cosine'), lty = c(2, 3), 
lwd = c(1.5, 1), bty = 'n')
title(main = 'Sine and Cosine functions')
```

```{r}
plot(temp$pretest ~ temp$posttest, data = df, main = 'plot 1')
plot(temp$pretest ~ temp$posttest, data = df, main = 'plot 2')
plot(temp$pretest ~ temp$posttest, data = df, main = 'plot 3')
plot(temp$pretest ~ temp$posttest, data = df, main = 'plot 4')
par(opt)
```

```{r}
plot(temp$pretest ~ temp$posttest, data = df, main = 'plot 1')
plot.new()
plot.new()
plot(temp$pretest ~ temp$posttest, data = df, main = 'plot 4')
par(opt)
```

### Simple customized Functions with multiple lines

```{r}
cummeadian = function(x) {
tmp = seq_along(x)
for(i in 1:length(tmp)) tmp[i] = median(x[1:i])
print(x)}
```

```{r}
cummeadian(temp$posttest)
```

### One-Line Functions

```{r}
log2 = function(x) log(x, base = 2)
log2
```

```{r}
log2(temp$posttest)
```

### Using Default Values in Functions

```{r}
manning = function(radius, gradient, coef=0.1125) (radius^(2/3)*gradient^0.5/coef)
```

```{r}
manning(radius = 1, gradient = 1/500)
```

### Chaining Functions Together with %>%, the Pipe Operator

```{r}
df %>% filter(df$school == 'ANKYI') %>% summary()
```

### Without the pipe operator, the preceding code would be written like this:

```{r}
summary(filter(df, school == 'ANKYI'))
```

### Creating a Scatter Plot

```{r}
plot(temp$pretest, temp$posttest)
```

### With ggplot2, you can get a similar result using the ggplot() function

```{r}
library(ggplot2)
```

```{r}
ggplot(temp, aes(x = temp$pretest, y = temp$posttest)) + geom_point()
```

### Creating a Line Graph

```{r}
plot(temp$pretest, type = 'l')
```

### With ggplot2

```{r}
ggplot(temp, aes(x = posttest, y = pretest)) + geom_line()
```

### Creating a Bar Graph

```{r}
ggplot(temp, aes(x = temp$posttest, y = temp$pretest)) + geom_col()
```

###  Creating a Histogram

```{r}
ggplot(temp, aes(x = temp$posttest)) + geom_histogram(binwidth = .5)
```

### Creating a Box Plot

```{r}
ggplot(temp, aes(x = temp$pretest, y = temp$posttest)) + geom_boxplot()
```

```{r}
ggplot(temp, aes(x = pretest, y = posttest)) + geom_col(fill = "lightblue", colour = "black")

```

### Grouping Bars Together

```{r}
ggplot(df, aes(x = school_setting, y = school, fill = posttest)) + geom_col(position = "dodge")
```

```{r}
ggplot(df, aes(x = school_type, y = school, fill = posttest)) + geom_col(position = "dodge")
```

```{r}
ggplot(df, aes(x = school_type, y = school, fill = posttest)) + geom_col(position = "dodge")
```

```{r}
ggplot(df, aes(x = teaching_method, y = school, fill = posttest)) + geom_col(position = "dodge")
```

```{r}
ggplot(df, aes(x = n_student, y = school, fill = posttest)) + geom_col(position = "dodge")
```

```{r}
ggplot(df, aes(x = gender, y = school, fill = posttest)) + geom_col(position = "dodge")
```

```{r}
ggplot(df, aes(x = lunch, y = school, fill = posttest)) + geom_col(position = "dodge")
```

```{r}
ggplot(df, aes(x = school_type, y = school, fill = school_setting)) + geom_col(position = "dodge")
```

```{r}
ggplot(temp, aes(x = gender, y = school_setting, fill = posttest)) + geom_col()
```

###  Making a Cleveland Dot Plot

```{r}
ggplot(df, aes(x = n_student, y = school)) + geom_point()
```

```{r}
ggplot(df, aes(x = n_student, y = school)) +
 geom_point(size = 3) + # Use a larger dot
 theme_bw() +
 theme(
 panel.grid.major.x = element_blank(),
 panel.grid.minor.x = element_blank(),
 panel.grid.major.y = element_line(colour = "grey60", linetype = "dashed")
 )
```

```{r}
ggplot(df, aes(x = n_student, y = school)) +
geom_segment(aes(yend = school), xend = 0, colour = "grey50") + 
  geom_point(size = 2) 
```

```{r}
ggplot(df,aes(x=school_setting, y=pretest, colour= gender)) + geom_point()
```

```{r}
ggplot(df,aes(x=school_setting, y=posttest, colour = school)) + geom_point()
```

```{r}
plot(df)
```

```{r}
ggplot(df, aes(x = df$school_setting, y = df$n_student)) +
 geom_point(size = 1.5)
```

```{r}
tt <- df %>%
 filter(df$school_type == 'Public') %>%
 select(school, n_student, pretest, posttest)
head(tt)
```

```{r}
tt = df %>% filter(df$gender == 'Female') %>% select(school, n_student, teaching_method, pretest, posttest)
head(tt)
```

```{r}
summary(tt)
```

```{r}
tt = df %>% filter(df$gender == 'Male') %>% select(school, n_student, teaching_method, pretest, posttest)
head(tt)
```

```{r}
summary(tt)
```

```{r}
tt = df %>% group_by(df$school) %>% summary()
tt
```